---
output: 
  pdf_document:
    toc: true
    number_sections: true
title: "HSTP Analysis"
author: "Tim Riley"
date: "`r format(Sys.time(), '%B %d, %Y')`"
header-includes:
  \usepackage[default,oldstyle,scale=0.95]{opensans}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE
)
#source(here::here("00_initialize.R"))
```

\pagebreak

# Introduction

This document presents findings from the analysis of newly-released ACS data for the HSTP update. All maps are saved in the /map subfolder of this repository.

# Topics

Tracts with more than 50% urban population in the 2010 census are excluded,

## Population Characteristics

This section covers general characteristics of the population in the rural areas of Region 6.

### Population Density

Mostly low density areas with a few small urban centers.

```{r population density map}
Region6TractPopulationDensityMap
```

\pagebreak

## Age Groups

This section breaks down the population by age groups.

### Under 18

Calculated using detailed table B01001, "SEX BY AGE".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_blw18))`.

Areas of McLean and Livingston counties have very high numbers of minors as proportion of the population.

```{r percent below 18 map}
Region6TractPercentBelow18Map
```

### Between 18 and 65

Calculated using detailed table B01001, "SEX BY AGE".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_18to65))`.

```{r percent 18 to 65 map}
Region6TractPercent18to65Map
```

### Over 65

Calculated using detailed table B01001, "SEX BY AGE".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_ovr65))`.

Possible significant correlation between lower population density and elderly population among Illinois Counties in general. This seems to hold in Region 6, where 20% or more of the population is over 65 in `r length(with(tractLayer, which(per_ovr65 > .2)))` of `r length(tractLayer)` tracts.

```{r percent over 65 map}
Region6TractPercentOver65Map
```

\pagebreak

## Disability

### Percent of adults with any disability

Calculated using detailed table B18101, "SEX BY AGE BY DISABILITY STATUS".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_dsblty))`.

```{r disability map}
Region6TractPercentDisabilityMap
```

\pagebreak

## Veterans

### Veterans as percent of all adults

Calculated using detailed table B21001, "SEX BY AGE BY VETERAN STATUS FOR THE CIVILIAN POPULATION 18 YEARS AND OVER".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_vet))`.

Rural areas of Livingston and Iroquois counties have proportionally high numbers of Veterans, and are somewhat far removed from the VA facilities in Kankakee and Bloomington.

```{r veterans map}
Region6TractPercentVeteranMap
```

\pagebreak

## Poverty

### Poverty Rate

Calculated using detailed table B17020, "POVERTY STATUS IN THE PAST 12 MONTHS BY AGE".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_blwpov))`.

Poverty in Region 6 is most highly concentrated in the small urban centers. Also Hopkins Park in Kankakee, historically an economically depressed area.

```{r poverty rate map}
Region6TractPercentBelowPovertyMap
```

### Households with elderly members on SNAP

Calculated using detailed table B22001, "RECEIPT OF FOOD STAMPS/SNAP IN THE PAST 12 MONTHS BY PRESENCE OF PEOPLE 60 YEARS AND OVER FOR HOUSEHOLDS".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_hhSo60))`.

Hopkins Park and parts of Pontiac.

```{r households on SNAP with persons over 60 map}
Region6TractPercentHouseholdSNAPOver60Map
```

\pagebreak

## Income

### Median Household Income

Calculated using detailed table B22008, "MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS (IN 2018 INFLATION-ADJUSTED DOLLARS) BY RECEIPT OF FOOD STAMPS/SNAP IN THE PAST 12 MONTHS".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$inc_medhh))`.


```{r median household income map}
Region6TractMedianHouseholdIncomeMap
```

### Income Per Capita

Calculated using detailed table B19301, "PER CAPITA INCOME IN THE PAST 12 MONTHS (IN 2018 INFLATION-ADJUSTED DOLLARS)".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$inc_percap))`.

Similar distribution to median household income

```{r per capita income map}
Region6TractIncomePerCapitaMap
```

### GINI

Calculated using detailed table B19083, "GINI INDEX OF INCOME INEQUALITY".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$gini))`.

Possible relationship to poverty

```{r gini coefficient map}
Region6TractGINICoefficientMap
```

\pagebreak

## Transportation

### Percent of Households with No Vehicle Available

Calculated using detailed table B08141, "MEANS OF TRANSPORTATION TO WORK BY VEHICLES AVAILABLE".

The average value for Illinois counties in `r acsYear` was `r percent(mean(illinoisCountyData$per_nocars))`.

Eastern Kankakee County, Pontiac in Livingston County, and Le Roy in McLean County stand out.

```{r percent no vehicle map}
Region6TractPercentNoCarMap
```